# ---
title: "Yusuf's Coursework 1"
author: "Yusuf Mohammad"
date: "2024-03-11"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("QMlogo.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:30%;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<h2><u>Student numbers in higher education - will the Trend continue?</u></h2>

In this project the data we will be exploring tells us the number of students in higher education on a 
yearly basis. The data has three categories, undergraduate, postgraduate and total, I will select the total
category from the dataset, the code to do so will be included in this report.

My goal with this project is to explore the trend behind the rising number of students in UK higher education
and to find out if this trend will continue and if so what can we expect going forwards. This area is of 
interest to universities who must adapt to the unforeseen demand of higher education and deal with the
prospect that the demand is not sustainable.

Lets load the data and take a look at it! Note we also filter for only the total category here and modify the dataframe such that the time is in column ds and the data in column y.
```{r fig.width = 10}
library(ggplot2)

# Load the dataset
data <- read.csv('students_by_study_type.csv')

# Filter for "Total" category
total_data <- data[data$Level.of.study == "Total",]

total_data$Academic.Year <- as.Date(paste0(substr(total_data$Academic.Year, 1, 4), "-01-01"))

#total_data = data.frame(total_data)

# Rename the columns
names(total_data)[names(total_data) == "Academic.Year"] <- "ds"
names(total_data)[names(total_data) == "Number"] <- "y"
total_data = subset(total_data, select = c(ds, y))
rownames(total_data) <- NULL

# Lets take a look at the times in our timeseries
print(total_data$ds)

# Create a line plot
p<- ggplot(total_data, aes(x = ds, y = y, group=1)) +
  geom_line() + 
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") + # Set x-axis breaks to every two years
  labs(title = "Total Number of Students per Academic Year",
       x = "Academic Year",
       y = "Number of Students")
p
```


```{r}
library(prophet)

model1 = prophet(total_data)

future = make_future_dataframe(model1, 2, freq = "year")
future
preds = predict(model1, future)
plot(model1, preds)
```
```{r}
plot_forecast_component(model1, preds, name = "trend")

```

```{r}
total_data
```
